\vspec{} and Friends: A suite of utilities to model spectroscopic phase curves of 3D exoplanet atmospheres in the presence of stellar variability.

### 2024-06-06 src/tex/ms.tex
Write a description of VSPEC overall

\newcommand{\teff}{$T_{\rm eff}$}
\definecolor{tedcommentcolor}{HTML}{e17701}
\newcommand{\TJ}[1]{\textcolor{tedcommentcolor}{#1}}

\shorttitle{\sc VSPEC}
\shortauthors{Johnson et al.}

@@

% Title
\title{The {\sc VSPEC} code: Spectroscopic phase curves of 3D exoplanet models in the presence of stellar variability}

@@

% Abstract with filler text
\begin{abstract}
    We present the Variable Star Phase Curve (\href{https://github.com/VSPEC-collab/VSPEC}{ \sc VSPEC}) code,
    a python package to simulate combined-light spectroscopic observations of exoplanets in the presence of stellar variability and inhomogeneity.
    {\sc VSPEC} uses the Planetary Spectrum Generator's Global Emission Spectra (PSG GlobES) application along with a custom-built
    stellar model based on an existing grid of stellar photosphere models to produce spectroscopic light curves of the planet-host system.
    VSPEC can be a useful tool for modeling observations of exoplanets in transiting geometries (primary transit, secondary eclipse) as well as orbital phase curve measurements,
    and is built in a modular and flexible configuration for easy adaptability to new stellar and planetary model inputs.
    We also describe associated codes. \TJ{These codes should be named here and described a little}
\end{abstract}

@@

In this paper we present present {\sc VSPEC}: Variable Star PhasE Curve\footnote{\url{https://github.com/VSPEC-collab/VSPEC}},
an open-source Python 3 package to simulate observations of exoplanet systems with variable host stars.
{\sc VSPEC} combines NASA's Planetary Spectrum Generator \citep[PSG,][]{villanueva2018} with a grid of
custom PHOENIX models \citep{husser2013} to produce multiwavelength phase curves of the system, accounting
for a 3D planetary atmosphere, transits and eclipses, and multiple sources of stellar variability.

In this paper we will describe the science behind {\sc VSPEC} and demonstrate examples of its use.
Section \ref{sec:star} describes the stellar model and the available sources of variability.
Section \ref{sec:pl_model} discusses modeling the planetary atmosphere, including PSG/GlobES and the
built-in GCM. In Section \ref{sec:examples} we provide examples of {\sc VSPEC} use cases and demonstrate
its value to the exoplanet community. Finally, in Section \ref{sec:conclusion} we will discuss the future of the code and issues it might have.

{Stellar Model \label{sec:star}}

The {\sc VSPEC} star model is designed modularly to allow
for both simple and complex behaviors. Currently, it
is represented by a rectangular grid of points on the stellar surface,
each assigned an effective temperature. At any given time, the model computes
the surface coverage fractions of each temperature visible to the observer, accounting for the
spherical geometry, limb darkening, and any occultation by a transiting planet.

@@

\subsection{Stellar Spectra \label{subsec:spectra}}
Once the surface coverage is calculated, a composite spectrum is computed from a grid of PHOENIX stellar spectra
\citep{husser2013} As of \texttt{VSPEC 0.1}, we have spectra between 2300 K and 3900 K, with steps of
100 K. Each spectrum has $\log{g} = 5$ and solar metalicity.
The raw spectra span from 0.1 to 20 \micron~ with $5 \times 10^{-6}$ \micron~ steps, however binned
versions are pre-computed for faster runtimes.

@@

{\sc VSPEC} also takes in parameters describing the population of spots, such as the mean of the (lognormal) peak area distribution
\citep{bogdan1988}, the equilibrium surface coverage, and the distribution style.

@@

Before simulating an observation, {\sc VSPEC} pre-computes a population of flares based on the frequency-energy relationship from \citet{gao2022}:
\begin{equation} \label{eq:flare_freq}
    \log{(f{\rm [day]})} = \beta + \alpha~\log{(E/{\rm [erg]})}
\end{equation}

@@

\subsection{Interfacing through {\sc pypsg}}
\ref{subsec:pypsg}
{\sc VSPEC} interfaces with PSG using the {\sc pypsg} package \TJ{link}, which is essentially a Python wrapper for the PSG API system.

@@

\subsection{3D Planet models}
{\sc VSPEC} allows the user to upload any desired GCM. However, for questions that focus on detectability,
the specific physics of the GCM are less important than properties like day/night temperature, surface pressure, or
the major absorbers in the atmosphere. To simplify these studies, {\sc VSPEC} has a built-in GCM that takes in just a
few parameters and produces a simple-yet-robust exoplanet climate. Simplicity allows the user to know everything about
the model and reduce surprises, but its basis in thermodynamics and robust energy balance make this model fit for publication-level science \TJ{is this really true? probably not}.

The {\sc VSPEC} GCM centers around a 2D temperature map of the planet's surface. Based on \citet{cowan2011}, this map
is constructed completely given the incident flux from the star (parameterized by the host \teff and radius
and the planet's semimajor axis and Bond albedo) and the unitless thermal redistribution efficiency $\epsilon$.
This efficiency is 0 for a planet that reradiates instantly and $\gg 1$ for an atmosphere that efficiently redistributes heat to its night side.
Figure \ref{fig:thermal_inertia} shows a temperature map example for an Earth-like planet with $\epsilon = 2\pi$.
These temperature maps are validated by the energy balance of the planet; {\sc VSPEC} computes the incident and emergent
flux across the planet and asserts that they agree.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/cowan_gcm.pdf}
    \caption{
        Example of a surface temperature map created by {\sc VSPEC}. The solar flux at Earth,
        a Bond albedo of 0.3, and $\epsilon = 2\pi$ were used in this case.
        Notice that the hottest point is offset from the sub-stellar point due to thermal inertia.
    }
    \script{cowan_gcm.py}
    \label{fig:thermal_inertia}
\end{figure}

@@

We will do three \texttt{VSPEC} simulations:
\begin{enumerate}
    \item Bare rock, no spots.
    \item 1 bar \ce{H2O} atmosphere, no spots.
    \item Bare rock, 2.5\% coverage by 2700 K spots
\end{enumerate}

@@

\subsection{Phase Curve}
Analysis of JWST MIRI-LRS phase curves of GJ 1214b by \citet{kempton2023} did not consider stellar
contamination because of the slow rotation rate of the host star \citep[approximately 1/80$^{\text{th}}$ the orbital frequency,][]{cloutier2021}.
In this example we use {\sc VSPEC} to demonstrate that this is a reasonable assumption.

We initialize {\sc VSPEC} to run a 41.0 hour observation with a cadence of 15 minutes. We use stellar and planetary parameters
from the NExScI Exoplanet Archive\footnote{\url{https://exoplanetarchive.ipac.caltech.edu/overview/GJ1214b}}.
The GCM used has a thermal inertia of $\epsilon = 6$ and a 1 bar \ce{CO2} atmosphere.

@@

After running the {\sc VSPEC} simulations, we bin into 0.5 $\mu m$ wavelength channels and normalize each channel to the two eclipses.
We assume the star behaves linearly between eclipse measurements and attribute any deviation to the planet.

@@

\subsection{MIRECLE Phase Curve}
In this example we will look at the best-case target for the Mid-IR Exoplanet CLimate Explorer \citep[MIRECLE,][]{mandell2022}
mission concept: Proxima Centauri b with no stellar variability. We will instead examine the expected noise produced by this observation
and assess the detectability of the planet in a single orbit. {\sc VSPEC} uses the noise models built into PSG, allowing native support for
noise due to the instrument, source, and astrophysical background.

We use \texttt{VSPEC}'s built-in MIRECLE instrument template, which includes a 2 m aperture and observes from 1-18 $\mu m$ with $R=50$.
We draw stellar and bulk planetary parameters from the NExScI Exoplanet Archive\footnote{\url{https://exoplanetarchive.ipac.caltech.edu/overview/proxima\%20cen}}
and assume $i=85^\circ$ and Earth density. The GCM we use has a Bond albedo of 0.3, $\epsilon=1.5$, $P_{\rm surf} = 1 \text{bar}$ and a 100 ppm \ce{CO2} atmosphere with a \ce{N2} background.
We ran the simulation for one orbit with an 24 hour cadence. Because this is a best-case scenario, we assumed perfect knowledge of the stellar spectrum.
We are left with the planetary spectrum and photon noise from the star.

>>>>

\newcommand{\teff}{$T_{\rm eff}$}
\definecolor{tedcommentcolor}{HTML}{e17701}
\newcommand{\TJ}[1]{\textcolor{tedcommentcolor}{#1}}
\newcommand{\vspec}[1]{\texttt{VSPEC}#1}
% \def\vspec#{\texttt{VSPEC}}

\shorttitle{\vspec{}}
\shortauthors{Johnson et al.}

@@

% Title
% \title{The \vspec{} code: Spectroscopic phase curves of 3D exoplanet models in the presence of stellar variability}
\title{\vspec{} and Friends: A suite of utilities to model spectroscopic phase curves of 3D exoplanet atmospheres in the presence of stellar variability.}

@@

% Abstract with filler text
\begin{abstract}
    We present the Variable Star Phase Curve (\href{https://github.com/VSPEC-collab/VSPEC} \vspec{}) code,
    a python package to simulate combined-light spectroscopic observations of exoplanets in the presence of stellar variability and inhomogeneity.
    \vspec{} uses the Planetary Spectrum Generator's Global Emission Spectra (PSG GlobES) application along with a custom-built
    stellar model based on an existing grid of stellar photosphere models to produce spectroscopic light curves of the planet-host system.
    \vspec{} can be a useful tool for modeling observations of exoplanets in transiting geometries (primary transit, secondary eclipse) as well as orbital phase curve measurements,
    and is built in a modular and flexible configuration for easy adaptability to new stellar and planetary model inputs. We additionally present
    a set of codes developed alongside \vspec{}, including the stellar surface model \texttt{vspec-vsm}, the stellar spectral grid interpolation code
    Gridpolator, and a Python interface for PSG \texttt{pypsg}.
\end{abstract}

@@

In this paper we present \vspec{}: Variable Star PhasE Curve\footnote{\url{https://github.com/VSPEC-collab/VSPEC}},
an open-source Python 3 package to simulate observations of exoplanet systems with variable host stars.
\vspec{} itself is merely an interface that pulls together a variable star surface model \texttt{vspec-vsm}\footnote{\url{https://github.com/VSPEC-collab/vspec-vsm}},
a stellar spectra grid interpolation code Gridpolator\footnote{\url{https://github.com/VSPEC-collab/GridPolator}}, and a Python interface for
the Planetary Spectrum Generator \citep[PSG,][]{villanueva2018}, \texttt{pypsg}. All of these codes together allow \vspec{} to account for
3D planetary atmospheres, time-resolved effects of both planet and star, many geometries including transit and eclipse, and realistic noise modeling.

\TJ{Change this later-->}
In this paper we will describe the science behind \vspec{} and demonstrate examples of its use.
Section \ref{sec:star} describes the stellar model and the available sources of variability.
Section \ref{sec:pl_model} discusses modeling the planetary atmosphere, including PSG/GlobES and the
built-in GCM. In Section \ref{sec:examples} we provide examples of \vspec{} use cases and demonstrate
its value to the exoplanet community. Finally, in Section \ref{sec:conclusion} we will discuss the future of the code and issues it might have.

\section{Using \vspec{}}
\label{sec:vspec}

The flow of \vspec{}'s code is broken into three parts:
\begin{enumerate}
    \item Read in configuration \\
    \item Compute planetary spectra \\
    \item Compute stellar spectra
\end{enumerate}

The main object that a user interacts with is the \texttt{ObservationModel}. This object can be initialized by passing it either a YAML file containing
all of the simulation information, or a \texttt{InternalParameters} object encoded with the same information. Upon initialization a local working directory
(by convention \texttt{.vspec}) is created to store simulation outputs and intermediate files.

The second step is initiated by calling the \texttt{ObservationModel.build\_planet()} method. This is where all the interaction with PSG takes place.
First, your 3D Global Circulation Model (GCM) is uploaded to PSG via the API's \texttt{set} method. Then, PSG is updated with information about the observation
that does not change with time, such as instrument parameters and system information. Before entering the main loop \vspec{} computes the geometry of the system
at each epoch, storing it for later. \vspec{} then iterates through each phase, making four calls to the PSG API:
\begin{enumerate}
    \item An \texttt{upd} request setting the geometric parameters for this epoch. \\
    \item An \texttt{all} request to retrieve the combined light spectrum. \\
    \item A second \texttt{upd} request to set the stellar flux to zero. \\
    \item A second \texttt{all} request to retrieve the thermal emission of the planet.
\end{enumerate}

Results from these calls are stored locally in the \texttt{.fits} format.

The last step is to replace the stellar spectra used by PSG with \vspec{}'s stellar model. First, this model is initialized based on the user-passed parameters.
One consideration made during this step is that a user may want a shorter cadence here than is necessary to resolve temporal features in the planetary spectrum,
so the geometry is recalculated and the planetary spectrum at each epoch is a linear interpolation of the results from the previous step. After this setup
\vspec{} enters the main loop, where it computes the composite stellar spectrum based on the stellar model and spectral grid. For each epoch it returns this stellar
spectrum as seen by the observer, the spectrum as seen by the planet, the reflected spectrum, the planetary thermal spectrum, the total combined spectrum, and the
total noise spectrum. The data at each epoch is stored in a \texttt{.fits} file.

The final outputs of \vspec{} should be easily read using Astropy's \texttt{QTable} object, but we have also included a class \texttt{PhaseAnalyzer} to automatically
parse all output files and conveniently access 1D spectra or lightcurves.

\section{Stellar Model}
\label{sec:star}

The \vspec{} star model is designed modularly to allow
for both simple and complex behaviors. Currently, it
is represented by a rectangular grid of points on the stellar surface,
each assigned an effective temperature. At any given time, the model computes
the surface coverage fractions of each temperature visible to the observer, accounting for the
spherical geometry, limb darkening, and any occultation by a transiting planet.

@@

\subsection{Stellar Spectra \label{subsec:spectra}}
Once the surface coverage is calculated, a composite spectrum is computed from a grid of PHOENIX stellar spectra
\citep{husser2013}.
As of \texttt{VSPEC 0.1} \TJ{fix!}, we have spectra between 2300 K and 3900 K, with steps of
100 K. Each spectrum has $\log{g} = 5$ and solar metalicity.
The raw spectra span from 0.1 to 20 \micron~ with $5 \times 10^{-6}$ \micron~ steps, however binned
versions are pre-computed for faster runtimes.

@@

\vspec{} also takes in parameters describing the population of spots, such as the mean of the (lognormal) peak area distribution
\citep{bogdan1988}, the equilibrium surface coverage, and the distribution style.

@@

Before simulating an observation, \vspec{} pre-computes a population of flares based on the frequency-energy relationship from \citet{gao2022}:
\begin{equation} \label{eq:flare_freq}
    \log{(f{\rm [day]})} = \beta + \alpha~\log{(E/{\rm [erg]})}
\end{equation}

@@

\subsection{Interfacing through {\sc pypsg}}
\ref{subsec:pypsg}
\vspec{} interfaces with PSG using the {\sc pypsg} package \TJ{link}, which is essentially a Python wrapper for the PSG API system.

@@

\subsection{3D Planet models}
\vspec{} allows the user to upload any desired GCM. However, for questions that focus on detectability,
the specific physics of the GCM are less important than properties like day/night temperature, surface pressure, or
the major absorbers in the atmosphere. To simplify these studies, \vspec{} has a built-in GCM that takes in just a
few parameters and produces a simple-yet-robust exoplanet climate. Simplicity allows the user to know everything about
the model and reduce surprises, but its basis in thermodynamics and robust energy balance make this model fit for publication-level science \TJ{is this really true? probably not}.

The \vspec{} GCM centers around a 2D temperature map of the planet's surface. Based on \citet{cowan2011}, this map
is constructed completely given the incident flux from the star (parameterized by the host \teff and radius
and the planet's semimajor axis and Bond albedo) and the unitless thermal redistribution efficiency $\epsilon$.
This efficiency is 0 for a planet that reradiates instantly and $\gg 1$ for an atmosphere that efficiently redistributes heat to its night side.
Figure \ref{fig:thermal_inertia} shows a temperature map example for an Earth-like planet with $\epsilon = 2\pi$.
These temperature maps are validated by the energy balance of the planet; \vspec{} computes the incident and emergent
flux across the planet and asserts that they agree.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/cowan_gcm.pdf}
    \caption{
        Example of a surface temperature map created by \vspec{}. The solar flux at Earth,
        a Bond albedo of 0.3, and $\epsilon = 2\pi$ were used in this case.
        Notice that the hottest point is offset from the sub-stellar point due to thermal inertia.
    }
    \script{cowan_gcm.py}
    \label{fig:thermal_inertia}
\end{figure}

@@

We will do three \vspec{} simulations:
\begin{enumerate}
    \item Bare rock, no spots.
    \item 1 bar \ce{H2O} atmosphere, no spots.
    \item Bare rock, 2.5\% coverage by 2700 K spots
\end{enumerate}

@@

\subsection{Phase Curve}
Analysis of JWST MIRI-LRS phase curves of GJ 1214b by \citet{kempton2023} did not consider stellar
contamination because of the slow rotation rate of the host star \citep[approximately 1/80$^{\text{th}}$ the orbital frequency,][]{cloutier2021}.
In this example we use \vspec{} to demonstrate that this is a reasonable assumption.

We initialize \vspec{} to run a 41.0 hour observation with a cadence of 15 minutes. We use stellar and planetary parameters
from the NExScI Exoplanet Archive\footnote{\url{https://exoplanetarchive.ipac.caltech.edu/overview/GJ1214b}}.
The GCM used has a thermal inertia of $\epsilon = 6$ and a 1 bar \ce{CO2} atmosphere.

@@

After running the \vspec{} simulations, we bin into 0.5 $\mu m$ wavelength channels and normalize each channel to the two eclipses.
We assume the star behaves linearly between eclipse measurements and attribute any deviation to the planet.

@@

\subsection{MIRECLE Phase Curve}
In this example we will look at the best-case target for the Mid-IR Exoplanet CLimate Explorer \citep[MIRECLE,][]{mandell2022}
mission concept: Proxima Centauri b with no stellar variability. We will instead examine the expected noise produced by this observation
and assess the detectability of the planet in a single orbit. \vspec{} uses the noise models built into PSG, allowing native support for
noise due to the instrument, source, and astrophysical background.

We use \vspec{}'s built-in MIRECLE instrument template, which includes a 2 m aperture and observes from 1-18 $\mu m$ with $R=50$.
We draw stellar and bulk planetary parameters from the NExScI Exoplanet Archive\footnote{\url{https://exoplanetarchive.ipac.caltech.edu/overview/proxima\%20cen}}
and assume $i=85^\circ$ and Earth density. The GCM we use has a Bond albedo of 0.3, $\epsilon=1.5$, $P_{\rm surf} = 1 \text{bar}$ and a 100 ppm \ce{CO2} atmosphere with a \ce{N2} background.
We ran the simulation for one orbit with an 24 hour cadence. Because this is a best-case scenario, we assumed perfect knowledge of the stellar spectrum.
We are left with the planetary spectrum and photon noise from the star.
